ms_supplementary

### 2018-01-29 man/ms_0_intro.Rmd
updating texts

@@

@@

`$NOTA$`: NDVI sirve para estimar la producción primaria neta. Existen diferentes estudio que han evaluado el efecto de la sequía sobre la producción primaria neta utilizando NDVI.

### Aims
`$IMPROVE$`

In this study we combined remote sensing information and dendroecological methods to evaluate the drought impacts in both greenees and growth of *Q. pyrenaica* forests in Sierra Nevada. Specifically,

@@

* and to explore differences in the resilience metrics between populations located in contrasting slopes within the rear edge of the distribution of this species.

Do northern and southern populations of Pyrenean oak forests differ in their resistance, resilience or recovery to extreme drought events?

## References

>>>>

```{r setup, echo=FALSE}
show_text <- FALSE
```

@@

## 
```{r echo=FALSE, results='asis', eval=show_text}
cat("

@@

`$NOTA$`: NDVI sirve para estimar la producción primaria neta. Existen diferentes estudio que han evaluado el efecto de la sequía sobre la producción primaria neta utilizando NDVI. 
")
```

Objetivos del trabajo

### Aims
In this study we combined remote sensing information and dendroecological methods to evaluate the drought impacts in both greenees and growth of *Q. pyrenaica* forests in Sierra Nevada. Specifically,

@@

* and to explore differences in the resilience metrics between populations located in contrasting slopes within the rear edge of the distribution of this species.Do northern and southern populations of Pyrenean oak forests differ in their resistance, resilience or recovery to extreme drought events?

### 2018-01-29 man/ms_1_material_methods.Rmd
updating texts

The years 2005 and 2012 were chosen as drought years for this study because they have been documented as two of the worst drought years particularly for the southern of the Iberian Peninsula [@GarciaHerrera2007; @Gouveia2015; @Trigo2013] and the were charecterized by extreme drogught (Appendix S1). Aunque sabemos que existen otros años de sequía extrema (e.g. 1995, 1981) [@VicenteSerrano2014], no se han elegido por una limitación metodológica, esto es, nuestro objetivo es analizar el efecto de la sequía y la resiliencia a los eventos de sequía utilizando información de satélite y de dendrocronología, y la información de satélite utilizada (ver mas adelante) solamente está disponible desde el año 2000 en adelante.

@@

After the filter out process, we built the annual EVI profile for each pixel and then computed the EVI's annual mean values and the EVI anomaly for each pixel for the period 2000 - 2016. EVI mean was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual (r = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88).

@@

### Tree sampling 
Sampling was carried during autumn of 2016. Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ; northern aspect) and Cáñar (CA; southern aspect) (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High. All the sites were oak monospecific and representatives of two of the three population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 dominant trees were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively. We assessed stand competition of target trees by recording distance, azimuth, dbh, species and total height of all neighboring living trees with dbh > 7.5 cm in a circular plot of 10 m radius. We computed two distance independent indices: density in $\mathrm{trees \cdot ha^{-1}}$, and basal area (BA) in $\mathrm{m^{2} \cdot ha^{-1}}$; and one distance dependent indices: size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$. We used non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] for search differences between sites for height, DBH and competence indices. When significant differences were observed, we run multiple comparisons using the Dunn's-test [@Dunn1964] with Bonferroni adjustment to correct significance values. These analysis were performed uisng the `PMCMR` package [@Pohlert2014].

>>>>

The years 2005 and 2012 were chosen as drought years for this study because they have been documented as two of the worst drought years particularly for the southern of the Iberian Peninsula [@GarciaHerrera2007; @Gouveia2015; @Trigo2013] and the were charecterized by extreme drogught (Figure S1). Aunque sabemos que existen otros años de sequía extrema (e.g. 1995, 1981) [@VicenteSerrano2014], no se han elegido por una limitación metodológica, esto es, nuestro objetivo es analizar el efecto de la sequía y la resiliencia a los eventos de sequía utilizando información de satélite y de dendrocronología, y la información de satélite utilizada (ver mas adelante) solamente está disponible desde el año 2000 en adelante.

@@

After the filter out process, we built the annual EVI profile for each pixel and then computed the EVI's annual mean values and the EVI anomaly for each pixel for the period 2000 - 2016. EVI mean was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual (r = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88). We explored the temporal trends of EVI at a pixel scale. We computed the nonparametric Mann–Kendall test (Kendall, 1975; Mann, 1945), using the `trend` package [@Thorsten2017]. We set 0.05 the alpha level for the test.

@@

### Tree sampling 
Sampling was carried during autumn of 2016. Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ; northern aspect) and Cáñar (CA; southern aspect) (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High. All the sites were oak monospecific and representatives of two of the three population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 dominant trees were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively. We assessed stand competition of target trees by recording distance, azimuth, dbh, species and total height of all neighboring living trees with dbh > 7.5 cm in a circular plot of 10 m radius. We computed two distance independent indices: *density* in $\mathrm{trees \cdot ha^{-1}}$, and *basal area* (BA) in $\mathrm{m^{2} \cdot ha^{-1}}$; and one distance dependent indices: size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$. We used non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] for search differences between sites for height, DBH and competence indices. When significant differences were observed, we run multiple comparisons using the Dunn's-test [@Dunn1964] with Bonferroni adjustment to correct significance values. These analysis were performed uisng the `PMCMR` package [@Pohlert2014].

### 2018-01-29 man/ms_2_results.Rmd
updating texts

Trends analysis showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant trends. The strongest trends were observed in southwestern populations (Figure S1). Vegetation greenness of *Q. pyrenaica* forests were lower during both the 2005 and 2012 years than the greenness observed for the reference period (Figure 2a). The lowest values for EVI standardized anomalies were recorded in 2005 being singnificantly lower (-2.285 ± 0.029) than 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Figure 2b). Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought (Figure 2b).

According to the standardized anomalies, *Q. pyrenaica* forests suffered a browning during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), however no changes were observed for the 2012 drought (Figure S2).

@@

## Responses to drougth: tree-growth (BAI)

Although no significant differences were observed in the resistance of the oak growth to both drought events (Table 3; Figure 3b), the 2005 drought reduced growth more than 2012 did (Rt = 0.721 and 0.819 respectively) (Table S2). Northern population showed resistance values lower than southern ones, especially for 2005 drought event where the growth was reduced to 44.5 % of the preceding period (Figures 3b, 4b).

@@

Oak forests recovered their growth more rapidly after 2012 than after 2005 staying at levels of Rc < 1 (Table S2). Norhtern populations showed signigicantly higher values of recovery than southern ones (Tables 3, S2), which even not reach recovery values of Rc = 1 (Figure 3b, 4b).

@@

Significantly higher values of resilience of growth were observerd for the 2012 drought event than for the 2005 one (Table S2; Figure 4b). We found no differences of resilience values for growth between northern and southern populations (Table 3). For 2005 drought event the resilience of northern population was lower than southern ones, but opposite pattern was found for the 2012 drought event, with northern population showing values greater than unity (Figure 3b).

# Chronology

The trees of the southern population were older and larger than northern ones, particularly for the high elevation site (Ca-High) which had bigger and taller trees than the other sites (Tables 1, 2). Competence indices were similar among sites (Table 1). The basal area increment (BAI) were different between sites (Figure 5). El sitio mas elevado (Ca-High) mostró mayores valores de BAI que los otros sitios, mostrando además una tendencia positiva a partir de 1970. Se observan diferencias en cuanto al crecimiento entre los sitios de las poblaciones del sur. Ca_Low se parece en crecimiento mas a SJ que a Ca_High.

and higher values of tree-growth (Table 2).

El sitio caH mostró además los mayores valores de AR(1) y de MS

El sitio del norte, mostró mayores valores de MS (Table 2). ¿esto sugiere que este sitio tiene una alta limitación climática?

Different growth trends were observed for each site (Figure 5). Ca-High (the highest elevation site) had higher values of tree growth than the other sites.

# Climate and tree

Tree-growth exhibited a significant response to water availability more than to the temperatures. Cumulative precipitation of the hydrological year (September of the previous year to August of the growth year) influenced positively the tree-growth for the both northern and southern populations (FIG ). Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population. Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth, specially for northern population. Spring Maximum temperature was the most significant limitation of the tree-growth for the southern population while maximum temperatures of the previous August did for the northern one. Conversely, mean, minimum and maximum temperatures of current September positively influenced the tree-growth in the northern population.

# Disturbance

@@

En cuanto a la señal a nivel de sitio, para el Norte, vemos que tanto las supresiones como las liberaciones se observan para mas del 50 % de los árboles. En el caso del sur, vemos que algunas supresiones se observan para mas del 50 % de los árboles, mientras que las liberaciones observadas en torno a los 2000 se observan para mas del 50 % de los árboles

notas para discusión

En ambos casos (evi y bai) la sequía de 2005 y de 2012 redujeron el greenness y el crecimiento. Además en ambos casos la sequía de 2005 redujo mas la variable, sin embargo, para el greenness si existen diferencias entre ambos eventos de sequía, mientras que para el crecimiento no. Por tanto podemos decir que se ha reducido el greennes pero para el crecimiento esto está mas amortiguado.

Al explorar la resistencia por sitio y drought se observa una curiosidad. La resistencia de los robledales de CaLow se mantiene casi igual. La de la cara norte (SJ) sufre mucho en 2005 y luego se recupera un poco. La de caH presenta menor resistencia en 2012 que en 2005 (y eso llama la atención)

## References

>>>>

Trends analysis showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant trends. The strongest trends were observed in southwestern populations (Figure S2). Vegetation greenness of *Q. pyrenaica* forests were lower during both the 2005 and 2012 years than the greenness observed for the reference period (Figure 2a). The lowest values for EVI standardized anomalies were recorded in 2005 being singnificantly lower (-2.285 ± 0.029) than 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Figure 2b). Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought (Figure 2b).

According to the standardized anomalies, *Q. pyrenaica* forests suffered a browning during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), however no changes were observed for the 2012 drought (Figure S3).

@@

## Growth trends
The trees of the southern population were older and larger than northern ones, particularly for the high elevation site (Ca-High) which had bigger and taller trees than the other sites (Tables 1, 2). Competence indices were similar among sites (Table 1). The basal area increment (BAI) were different between sites (Figure 5). El sitio mas elevado (Ca-High) mostró mayores valores de BAI que los otros sitios, mostrando además una tendencia positiva a partir de 1970. Se observan diferencias en cuanto al crecimiento entre los sitios de las poblaciones del sur. Ca_Low se parece en crecimiento mas a SJ que a Ca_High.

Los tres sitios mostraron valores de ring width indices bajos para el año 1995 (Figure 6). El sitio SJ mostró valores muy bajos de ring width indices en 1995, 2005, y 2012. Sin embargo esto no ocurrió igual en los sitios del sur (caL y CaH) donde si encontramos valores bajos de RWI para 1995, pero no para 2005 y 2012.

Sorprende que en los sitios del sur (caH y caL), en los últimos años ha habido crecimientos muy grandes. Por ejemplo, en 2010 se observó uno de los mayores valores de RWI de toda la cronología tanto en caH como en caL. Algo similar ocurrió en sj para los años 2003 y 2013.

## Climate and tree
Tree-growth exhibited a significant greater response to water availability more than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge). Cumulative precipitation of the hydrological year was the climatic variable with a higher relationship with growth and influenced positively the tree-growth for the both northern and southern populations (Figure 7a). Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population. Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (r > 0.6) (Figure 7b), specially for northern population, which can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014]. Temperatures were much less related to growth than moisture variables. Yet spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while maximum temperatures of the previous August were the only negative significant for the northern one (Figures 7c and 7d). Surprisingly minimum and maximum temperatures of current September positively influenced the tree-growth in the northern population.

## Disturbances
The analysis of growth changes revealed differences between sites (Figures 8 and 9). En el sitio del norte osbervamos dos eventos de liberación importantes (GC > 50 %), que ocurren en mas de la mitad de los árboles muestreados (Figures 8 and 9). El primero de ellos desde 1940 hasta 1950, mientras que el segundo entre 1994 y 2001. Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??). Estos periodos se alternan con periodos de supresión. En los sitios del sur (caH y caL), sin embargo se observan algunos episodios débiles de supresión, aunque estos no se observan para los últimos años, de hecho incluso se observa una débil liberación (Figure 8) afectando a mas de la mitad de los árboles muestreados (Figure 9).

## Responses to drougth: tree-growth

Although no significant differences were observed in the resistance of oak growth between the two drought events (Table 3), the 2005 drought reduced growth more than that of 2012 (Rt = 0.721 and 0.819 respectively) (Table S2). The northern site, which is under a drier cliamte, showed resistance values lower than those of the southern site, especially for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figures 3b, 4b).

@@

Oak forests recovered their growth more rapidly after 2012 than after 2005 staying at levels of Rc < 1 (Table S2). Northern populations showed signigicantly higher values of recovery than southern ones (Tables 3 and S2), which even not reach recovery values of Rc = 1 (Figure 3b, 4b).

@@

Significantly higher values of resilience of growth were observerd for the 2012 drought event than for the 2005 one (Table S2; Figure 4b). We found no differences of resilience values for growth between northern and southern populations (Table 3). For 2005 drought event the resilience of northern population was lower than that of southern ones, but opposite pattern was found for the 2012 drought event, with northern population showing values greater than unity (Figure 3b).

```{r echo=FALSE, results='asis', eval=show_text}
cat("

@@

En cuanto a la señal a nivel de sitio, para el Norte, vemos que tanto las supresiones como las liberaciones se observan para mas del 50 % de los árboles. En el caso del sur, vemos que algunas supresiones se observan para mas del 50 % de los árboles, mientras que las liberaciones observadas en torno a los 2000 se observan para mas del 50 % de los árboles 
")
```

### 2018-01-29 man/ms_3_discussion.Rmd
updating texts

>>>>

and higher values of tree-growth (Table 2).

El sitio caH mostró además los mayores valores de AR(1) y de MS

El sitio del norte, mostró mayores valores de MS (Table 2). ¿esto sugiere que este sitio tiene una alta limitación climática?

Different growth trends were observed for each site (Figure 5). Ca-High (the highest elevation site) had higher values of tree growth than the other sites.

notas para discusión

En ambos casos (evi y bai) la sequía de 2005 y de 2012 redujeron el greenness y el crecimiento. Además en ambos casos la sequía de 2005 redujo mas la variable, sin embargo, para el greenness si existen diferencias entre ambos eventos de sequía, mientras que para el crecimiento no. Por tanto podemos decir que se ha reducido el greennes pero para el crecimiento esto está mas amortiguado.

Al explorar la resistencia por sitio y drought se observa una curiosidad. La resistencia de los robledales de CaLow se mantiene casi igual. La de la cara norte (SJ) sufre mucho en 2005 y luego se recupera un poco. La de caH presenta menor resistencia en 2012 que en 2005 (y eso llama la atención)

### 2018-01-29 man/ms_figures.Rmd
updating texts

```{r}
cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
year_cutoffs <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_cutoffs.csv"), header = TRUE, sep = ',')
year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')

label_site <- c("0_sj" = "Northern (SJ)",
                "1_caH" = "Southern High (CaH)",
                "2_caL" = "Southern Low (CaL)")

>>>>

```{r}
cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
year_cutoffs <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_cutoffs.csv"), header = TRUE, sep = ',')
year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')
```

```{r}
# top10  
top10pos50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
top10neg50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
top10pos <- cro_rwi %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
top10neg <- cro_rwi %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
```

```{r}
label_site <- c("0_sj" = "Northern (SJ)",
                "1_caH" = "Southern High (CaH)",
                "2_caL" = "Southern Low (CaL)")

### 2018-01-29 man/ms_supp.Rmd
updating texts

```{r}
library("tidyverse")
library("here")
library("dplyr")
library("ggrepel")
library("scales")
library("grid")
library("gridExtra")
library("patchwork")
library("ggcorrplot")
```

@@

**Figure S1**. Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016. Negative values (*red* values) indicate negative trends and positive values (*green* values) indicate positive trends.

@@

**Figure S2.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.

>>>>

```{r}
library("tidyverse")
library("here")
library("dplyr")
library("ggrepel")
library("scales")
library("grid")
library("gridExtra")
library("patchwork")
library("ggcorrplot")
library("stringr")
```

@@

**Figure S1** Acumulated monthly precipitation during the hydrological year 2004-2005 (blue line) and 2011-2012 (red line). The boxplot representing the average from 1940-2015 period. Data from meteorological station Granada, Base Aérea (National Spanish Meteorological Services (AEMET)).

```{r}
ltdf <- read.table(file=here("/data_raw/meteo", "long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')

source(here("/script/R/", "hydro_year.R"))

ltdf <- ltdf %>% 
  dplyr::select(cod = INDICATIVO, VARIABLE, FECHA, VALOR) %>% 
  dplyr::mutate(fecha = as.Date(FECHA, format='%d/%m/%Y'),
         month = lubridate::month(fecha),
         hmonth = ifelse(month <= 8, month + 4, month -8), 
         hyear = hydro_year(fecha),
         hyear_f = paste0(hyear-1,'-', hyear))

# We used only the 5514 station (the most long term)
# Filter by station and prec 
# Filter from 1950 
st5514 <- ltdf %>% 
  filter(VARIABLE == 'PM1') %>% 
  filter(cod == '5514') %>% 
  filter(hyear >= 1950) %>% 
  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR)

st5514_pro <- st5514 %>% 
  group_by(hyear_f, hmonth) %>% 
  dplyr::summarise(pre = sum(pre)) %>%
  dplyr::mutate(csum = cumsum(pre)) %>% as.data.frame()

st5514_pro  %>% 
  filter(!(hyear_f %in% c('2004-2005', '2011-2012'))) %>% 
  ggplot(aes(x=as.factor(hmonth), y=csum)) + 
  geom_boxplot(aes(x=as.factor(hmonth))) +  
  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
  geom_line(data=subset(st5514_pro, hyear_f == '2004-2005'),
            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
  geom_line(data=subset(st5514_pro, hyear_f == '2011-2012'),
            aes(x=as.factor(hmonth), y=csum, group=1),
            colour='red') +
  theme_bw() + 
  theme(strip.background = element_rect(fill = NA, color = "black")) + 
  ylab('Precipitation (mm) acumulated') + xlab('') + 
  ggtitle('Granada Base Aerea, 1950-2015') + 
  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
  annotate("text", label = "2004-2005", x = 2, y = 500, size = 6, colour = "blue") + 
  annotate("text", label = "2011-2012", x = 2, y = 550, size = 6, colour = "red") + 
  annotate("text", label = "mean 1950-2015", x = 3, y = 600, size = 6, colour = "black")
```

```{r, eval=FALSE, echo=FALSE}
# Get file names
myfiles <- list.files(path = here("/data_raw/spei_all/"), pattern = "\\.csv$")

mydf <- data.frame()

# Loop to read files and prepare data
for (j in myfiles){ 
  aux <- read.csv(file=paste0(here("/data_raw/spei_all/"), j), header = TRUE, sep = ',')

  # Remove csv and get name 
  name_aux <- str_replace(j, ".csv", "")

  # Get lat long
  latlong <- str_replace(name_aux, "SPEI_", "")

  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]

  aux$lat <- mylat
  aux$long <- mylong

  # Split time 
  # Get year 
  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))

  # Get month 
  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")

  # Variable month in number 
  aux$months <- match(aux$month, month.abb)

  # assign(name_aux, aux)

  mydf <- rbind(mydf, aux)
}

spei <- mydf %>% select(SPEI_12, year, months) %>% 
  group_by(year,months) %>% 
  summarise(spei6 = mean(SPEI_12),
            spei6_sd = sd(SPEI_12)) %>% 
  mutate(date = lubridate::ymd(paste(year, months, "01")))

spei <- spei %>% 
  mutate(sign = case_when(
    spei6 > 0 ~ "pos",
    TRUE ~ "neg"))

limits <- as.Date(c("19549-01-01", "2015-12-01"))
spei %>% 
  ggplot(aes(x=date, y=spei6, fill=sign)) +  
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
  theme_bw()  + 
  labs(title= 'Drought evolution for Sierra Nevada',
    x= '',
    y='SPEI') +
  theme(strip.background = element_rect(fill = "white"),
        legend.position = "none") +
  scale_x_date(date_breaks = "5 year", date_labels = "%Y", limits=limits) 
```

**Figure S2**. Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016. Negative values (*red* values) indicate negative trends and positive values (*green* values) indicate positive trends.

@@

**Figure S3.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
